Promise and peril: Agnostic identification methods\\ for detecting differential item functioning

### 2020-03-20 paper/paper_2a_anchor_items.Rmd
got another draft to ben

The unsolved problem that we address here is how to link groups in the typical setting in which no a prior assumptions — about either relative group ability or which items might have DIF — can be made. That is, we're interested in agnostic identification methods (hereafter referred to as "AI methods") so that LRTs can be used to test items for DIF. For simplicity, we focus on the Rasch model to isolate the fundamental issues in DIF detection. And, in search of a coherent framework, we sometimes edit names of existing methods. We recognize that others have done the same [e.g. @kopf2015framework], and that we risk contributing to a proliferation of names.

In this section, we summarize existing AI methods and propose a variety of extensions. We use a simple demonstration data to illustrate how these methods work: 10,000 reference group students and 10,000 focal group students taking an eight-item test. Target ability is simulated ${\theta_i}^{\text{ref}} \sim N(0,1)$ for students in the reference group and ${\theta_i}^{\text{foc}} \sim N(-1,1)$ for students in the focal group. Nuisance ability is set to ${\eta_i}^{\text{ref}} = 0$ and ${\eta_i}^{\text{foc}} = -1$. The slope on target ability is set to $a_{1j} = 1$ for all items. The slope on nuisance ability is set to $a_{2j} = 0.5$ for the last three items (the items with DIF) and $a_{2j} = 0$ otherwise. We can also, of course, describe these conditions from the DIF-as-item-property view where there is no nuisance ability. Instead, ${b_j}^{\text{foc}} = {b_j}^{\text{ref}}$ for the first five items and ${b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.5$ for the last three items. To be sure, this data is used only for demonstrative purposes. We compare the performance of the AI methods in a more realistic scenario in the [the simulation study](#simstudy).

@@

We think so and, in response, propose the "equal means, multiple imputation logit graph" (EM-MILG), which presents information about potential DIF to the analyst. EM-MILG begins by fitting a unidimensional Rasch model to the data that is identified by setting $\mu^\text{foc} = 0$. Recall that $\mu^\text{ref} = 0$, so the assumption is that the groups have equal mean ability. The result is that all differences in performance — either from group ability differences or DIF — manifest in the item easiness difference parameter $\tilde{d_j}$. The tilde is used to denote parameters estimated with $\mu^\text{foc}$ arbitrarily set to $0$.

@@

out_3_biased_items$AOAA_OAT_final_dif[[1]] %>% 
    mutate(difference_in_easy = difference_in_easy - difference_in_easy[anchor][1]) %>% 
    ggplot(aes(x = difference_in_easy, y = item)) +
    ggridges::geom_density_ridges(
        data = 
            out_3_biased_items$AOAA_OAT_mod[[1]] %>% 
            mod_to_draws_df(n = 1000) %>% 
            select(-run) %>%
            gather(var, val) %>% 
            filter(val != 0),
        aes(x = val, y = var)
    ) +
    geom_point() +
    labs(
        x = latex2exp::TeX("$\\tilde{d_j} = \\tilde{b_j}^{ref} - \\tilde{b_j}^{foc}$"),
        y = ""
    )
```

In general, one of our key concerns with traditional AI methods is that they can lull the analyst into a false sense of security. Too often, the analyst chooses a method, implements it, and then proceeds as if the method certainly resulted in a model that reflects some objective reality. A EM-MILG combats this concern by presenting all information clearly to the analyst. In the previous example, the analyst may be wary of their results, having seen how arbitrary it was to conclude that the first five and not the last three items are unbiased. Even when other AI methods are used, the analyst can use the EM-MILG as the first step in order to give them a sense of their item response data. And, of course, the MILG can be used to visualize DIF regardless of how the model is identified. We now move on to more traditional AI methods.

@@

One way to attempt to counter artificial DIF is with purification through iterative anchor selection. For example, @drasgow1987study started with AOAA, removed items displaying DIF from the anchor set, then repeated the process iteratively — with items that have been removed from the anchor set allowed to have free parameters across groups in both the baseline and flexible model — until no more items tested positively. @kopf2015framework named this technique Iterative-backward-AOAA with "backward" (as in reverse) referring to beginning with the assumption that all items are DIF-free. We find it clearer to refer to this method as all-others-as-anchors-all-significiant (AOAA-AS). Appending all-significant indicates that anchor selection is made iteratively with all items that test positive for DIF being removed from the anchor set. AOAA-AS is implemented in the mirt R package, and is called by passing scheme = "drop_sequential" to the DIF function.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been explicitly proposed. AOAA-OAT is inspired by @hagquist2017recent, who, in general, assert that "items showing DIF initially should not be resolved simultaneously but sequentially" [p. 6]. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the $\chi^2$ test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only one item in each iteration as opposed to all items that test positive for DIF. As a result, we believe that AOAA-OAT is less likely to be "fooled" by artifical DIF. Note that AOAA-OAT is not currently implemented in the R package mirt.

@@

@bechger2015statistical proposed selecting anchor items by identifying clusters of items that function similarly and then choosing one of those clusters to be the "anchor cluster." They pointed out that one way around the unidentifiability issue is to consider only relative item parameters. For each group, the relative easinesses for each pair of items can be stored in the matrix $\mathbf{R}^{\text{ref}}$ with entries ${R_{xy}}^{\text{ref}} = {b_x}^{\text{ref}} - {b_y}^{\text{ref}}$, where $x$ and $y$ both index the items. The ultimate matrix of interest is $\Delta \mathbf{R} \equiv \mathbf{R}^{\text{ref}}-\mathbf{R}^{\text{foc}}$ which is the "differences between groups in the pairwise differences in (easiness) between items" [p. 323].

>>>>

The unsolved problem that we address here is how to link groups in the typical setting in which no a prior assumptions — about either relative group ability or which items might have DIF — can be made. That is, we're interested in agnostic identification methods (hereafter referred to as "AI methods") so that LRTs can be used to test items for DIF. For simplicity, we focus on the Rasch model to isolate the fundamental issues in DIF detection. And, in search of a coherent framework, we sometimes edit names of existing methods. We recognize that others have done the same [e.g., @kopf2015framework], and that we risk contributing to a proliferation of names.

In this section, we summarize existing AI methods and propose a variety of extensions. We use simple demonstration data to illustrate how these methods work: 10,000 reference group students and 10,000 focal group students taking an eight-item test. Target ability is simulated ${\theta_i}^{\text{ref}} \sim N(0,1)$ for students in the reference group and ${\theta_i}^{\text{foc}} \sim N(-1,1)$ for students in the focal group. Nuisance ability is set to ${\eta_i}^{\text{ref}} = 0$ and ${\eta_i}^{\text{foc}} = -1$. The slope on target ability is set to $a_{1j} = 1$ for all items. The slope on nuisance ability is set to $a_{2j} = 0.5$ for the last three items (the items with DIF) and $a_{2j} = 0$ otherwise. We can also, of course, describe these conditions from the DIF-as-item-property view where there is no nuisance ability. Instead, ${b_j}^{\text{foc}} = {b_j}^{\text{ref}}$ for the first five items and ${b_j}^{\text{foc}} = {b_j}^{\text{ref}} - 0.5$ for the last three items. To be sure, this data is used only for demonstrative purposes. We compare the performance of the AI methods in a more realistic scenario in the [the simulation study](#simstudy).

@@

We think so and, in response, propose the "equal means, multiple imputation logit graph" (EM-MILG), which presents information about potential DIF to the analyst. EM-MILG begins by fitting a unidimensional Rasch model to the data that is identified by arbitrarily setting $\mu^\text{foc} = 0$. Recall that $\mu^\text{ref} = 0$, so this implies that the groups have equal mean ability. The result is that all differences in performance — either from group ability differences or DIF — manifest in the item easiness difference parameter $\tilde{d_j}$. The tilde (e.g., $\tilde{d_j}$) is used to denote parameters estimated with $\mu^\text{foc}$ set to $0$ — as compared to the hat (e.g., $\hat{d_j}$), which is used for parameter estimates from the final model.

@@

out_3_biased_items$AOAA_OAT_final_dif[[1]] %>% 
    mutate(difference_in_easy = difference_in_easy - difference_in_easy[anchor][1]) %>% 
    ggplot(aes(x = difference_in_easy, y = item)) +
    ggridges::geom_density_ridges(
        data = 
            out_3_biased_items$AOAA_OAT_mod[[1]] %>% 
            mod_to_draws_df(n = 1000) %>% 
            select(-run) %>%
            gather(var, val) %>% 
            filter(val != 0),
        aes(x = val, y = var)
    ) +
    geom_point() +
    labs(
        x = latex2exp::TeX("$\\hat{d_j} = \\hat{b_j}^{ref} - \\hat{b_j}^{foc}$"),
        y = ""
    )
```

In general, one of our key concerns with traditional AI methods is that they can lull the analyst into a false sense of security. Too often, the analyst chooses a method, implements it, and then proceeds as if the method certainly resulted in a model that reflects some objective reality. A EM-MILG combats this concern by presenting all information clearly to the analyst. In the previous example, the analyst may be wary of their results, having seen how arbitrary it was to conclude that the first five and not the last three items are unbiased. Even when other AI methods are used, the analyst can use the EM-MILG as the first step in order to give them a sense of their item response data. And, of course, the MILG can be used to visualize DIF regardless of how the model is identified. We now move on to more traditional AI methods, which don't require judgment from the analyst.

@@

One way to attempt to counter artificial DIF is with purification through iterative anchor selection. For example, @drasgow1987study started with AOAA, removed items displaying DIF from the anchor set, then repeated the process iteratively — with items that have been removed from the anchor set allowed to have free parameters across groups in both the baseline and flexible model — until no more items tested positively. @kopf2015framework named this technique Iterative-backward-AOAA with "backward" (as in reverse) referring to beginning with the assumption that all items are DIF-free. We find it clearer to refer to this method as all-others-as-anchors-all-significiant (AOAA-AS). Appending "all-significant" indicates that anchor selection is made iteratively with all items that test positive for DIF being removed from the anchor set. AOAA-AS is implemented in the mirt R package, and is called by passing scheme = "drop_sequential" to the DIF function.

@@

We propose an extension of these methods, all-others-as-anchors-one-at-a-time (AOAA-OAT), which, to our knowledge (and surprise), has not previously been explicitly proposed. AOAA-OAT is inspired by @hagquist2017recent, who, in general, assert that "items showing DIF initially should not be resolved simultaneously but sequentially" [p. 6]. Like AOAA-AS, AOAA-OAT starts with AOAA, but only the single item exhibiting the most DIF, based on the $\chi^2$ test statistic, is removed from the anchor set. The process continues iteratively until no new items display DIF. AOAA-OAT and AOAA-AS are similar in that they are both iterative; the difference is that AOAA-OAT takes the more conservative approach of removing only *one* item in each iteration as opposed to *all* items that test positive for DIF. As a result, we believe that AOAA-OAT is less likely to be "fooled" by artifical DIF. AOAA-OAT is not currently implemented in the R package mirt.

@@

@bechger2015statistical proposed selecting anchor items by identifying clusters of items that function similarly and then choosing one of those clusters to be the "anchor cluster." They pointed out that one way around the unidentifiability issue is to consider only relative item parameters. For each group, the relative easinesses for each pair of items can be stored in the matrix $\mathbf{R}^{\text{ref}}$ with entries ${R_{xy}}^{\text{ref}} = {b_x}^{\text{ref}} - {b_y}^{\text{ref}}$, where $x$ and $y$ both index the items. To be sure, $\mathbf{R}^{\text{ref}}$ is a square with the number of rows and columns equal to the number of items. The ultimate matrix of interest is $\Delta \mathbf{R} \equiv \mathbf{R}^{\text{ref}}-\mathbf{R}^{\text{foc}}$ which is the "differences between groups in the pairwise differences in (easiness) between items" [p. 323].

### 2020-03-20 paper/paper_2b_anchor_points.Rmd
got another draft to ben

$\mu^{\star\text{foc}}$ is selected by maximizing the Gini index (thus the abbreviation MAXGI). The intuition and assumption is that the items without DIF are the most homogeneous. Denoting a function that calculates the Gini index from a vector of non-negative elements as  $G(\mathbf{x})$, MGI sets <!-- its weird to arg max over a parameter we already have a definition for - maybe switch to t or something -->
\begin{align}
\mu^{\star\text{foc}} = \argmax_{\mu^\text{foc}} G(|\mu^\text{foc} + \tilde{\mathbf{d}}|)
\end{align}
where $\mu^\text{foc} \in (-\infty, \infty)$ and $\mu^\text{foc}$ is added to each element of $\mathbf{d}$.

@@

For our demonstration data simulation, Figure \ref{fig:ginipath} shows the gini coefficient at a variety of possible $\mu^\text{foc}$ values. The result of MAXGI is $\mu^{\star\text{foc}} = `r round(gini_results$gini_anchor_points, 2)`$, which is extraordinarily close to the data-generating value of -1.

@@

To adapt Raju's area method into an AI method, we propose a new method, which we call "minimizing the area between curves" (MINBC). To understand MINBC, imagine a scenario in which the data-generating process is $\mu^{\text{foc}} = \mu^{\text{ref}}$ and $d_j = 0 \forall j$, so that there is no DIF. The fundamental identification problem is that there are an infinite number of models with the same likelihood from which to choose. For example, we could correctly assume that the focal group has the same ability as the reference group and fix $\mu^{\star\text{foc}} = 0$. The model would then estimate $\hat d_j \approx 0 \forall j$, and we would correctly conclude the groups have the same ability and there is no DIF. Alternatively, we could assume that the focal group has $\mu^{\star\text{foc}} = 3$. The model would then compensate by finding $\hat d_j \approx -3 \forall j$, and we would incorrectly conclude that the focal group is high ability, but every item contains DIF against them. Both of these models have the same likelihood, so how should one choose which model to believe? MINBC chooses the model with the least amount of total DIF, as measured by the total weighted area between the item characteristic curves. As a result, the likelihood tie is broken by preferring to explain differences in performance across groups by ability differences (as opposed to DIF).

>>>>

$\mu^{\star\text{foc}}$ is selected by maximizing the Gini index (thus the abbreviation MAXGI). The intuition and assumption is that the anchor point should prioritize the majority of items having little to no DIF. Denoting a function that calculates the Gini index from a vector of non-negative elements as  $G(\mathbf{x})$, MGI sets <!-- its weird to arg max over a parameter we already have a definition for - maybe switch to t or something -->
\begin{align}
\mu^{\star\text{foc}} = \argmax_{\mu^\text{foc}} G(|\mu^\text{foc} + \tilde{\mathbf{d}}|)
\end{align}
where $\mu^\text{foc} \in (-\infty, \infty)$ and $\mu^\text{foc}$ is added to each element of $\tilde{\mathbf{d}}$.

@@

For our demonstration data simulation, Figure \ref{fig:ginipath} shows the gini coefficient at a variety of possible $\mu^\text{foc}$ values. The result of MAXGI is $\mu^{\star\text{foc}} = `r round(gini_results$gini_anchor_points, 2)`$, which is extraordinarily close to the data-generating value of -1. Moreover, there is a local maximum at $\mu^{\text{foc}} \approx 1.5$, which corresponds to the items with DIF. This illustrates that — as @strobl2018anchor point out — the search path is informative.

@@

To adapt Raju's area method into an AI method, we propose a new method, which we call "minimizing the area between curves" (MINBC). To understand MINBC, imagine a scenario in which the data-generating process is $\mu^{\text{foc}} = \mu^{\text{ref}}$ and $d_j = 0 \forall j$, so that the groups have equal ability and there is no DIF. The fundamental identification problem is that there are an infinite number of models with the same likelihood from which to choose. For example, we could correctly assume that the focal group has the same ability as the reference group and fix $\mu^{\star\text{foc}} = 0$. The model would then estimate $\hat d_j \approx 0 \forall j$, and we would correctly conclude the groups have the same ability and there is no DIF. Alternatively, we could assume that the focal group has $\mu^{\star\text{foc}} = 3$. The model would then compensate by finding $\hat d_j \approx -3 \forall j$, and we would incorrectly conclude that the focal group is high ability, but every item contains DIF against them. Both of these models have the same likelihood, so how should one choose which model to believe? MINBC chooses the model with the least amount of total DIF, as measured by the total weighted area between the item characteristic curves. As a result, the likelihood tie is broken by preferring to explain differences in performance across groups by ability differences (as opposed to DIF).

### 2020-03-20 paper/paper_3_sim.Rmd
got another draft to ben

In our simulation study, it was critical that student ability was drawn in a realistic way similar to @walker2017using. However, we don’t believe that a compensatory model is realistic in describing a math test where some items depend on verbal ability. For example, it’s hard to imagine that a student without the verbal ability to parse a word problem could fully compensate by having a higher math ability. Accordingly, we generated item responses using a simplified version of Sympson’s -@sympson1978model noncompensatory item response model in which
\begin{align}
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{2j}) = \sigma(\theta_i) \cdot \sigma(a_{2j}\eta_i)
\end{align}
where, as before, $\theta_i$ is target ability, $\eta_i$ is nuisance ability, and $\sigma(a_{2j})$ is the item's loading on nuisance ability [@demars2016partially].

@@

The test always has 12 items, but we varied the number of items with DIF from two to six. For items without DIF, $a_{2j} = \infty$ so that the model reduces to $\text{Pr}(y_{ij} = 1 | \theta_i) = \sigma(\theta_i)$. For items with DIF, $a_{2j}$ is calculated based on Ackerman's -@ackerman1994using angle equation as described in @walker2017using:
\begin{align}
\angle_j = \arccos \dfrac{a_{1j}^2}{a_{1j}^2 + a_{2j}^2}.
\end{align}
An item's angle measures the relative loading of the item on the two dimensions. For example, an angle of 45$^\circ$ indicates that the item loads equally on the target ability and nuisance ability. Our simple noncompensatory model has $a_{1j} = 1$ for all items so the angle equation reduces to
\begin{align}
\angle_j = \arccos \dfrac{1}{1 + a_{2j}^2}.
\end{align}
We are interested in specifying the angle of an item, so the relevant equation becomes
\begin{align}
a_{2j} = \sqrt{\dfrac{1 - \cos(\angle_j)^2}{\cos(\angle_j)^2}}.
\end{align}
For DIF items, we set $a_{2j}$ based on angles with equal intervals between 20$^\circ$ and 60$^\circ$. For example, for a test with three DIF items the angles are 20$^\circ$, 40$^\circ$, and 60$^\circ$.

@@

```{r simemmlg, fig.cap = 'For a 20 item test, the relationship between target ability and probability of correct response with nuisance abilities fixed to group means.', out.width="70%", warning = FALSE, message = FALSE}
out$bigsim[[5]]$intuitive_mod[[1]] %>%
    mod_intuitive_to_draws_df() %>%
    draws_df_to_logit_plot() +
    labs(
          x = latex2exp::TeX("$\\tilde{d_j} = \\tilde{b_j}^{ref} - \\tilde{b_j}^{foc}$")
      )
```

@@

An effective AI method should lead to a final model that accurately estimates the difference between the reference group's mean target ability and the focal group's mean target ability. We refer to this quantity as the achievement gap. Recall that all models set $\mu_\theta^\text{ref} = 0$, so the achievement gap reduces to $\mu_\theta^\text{ref}$. The data-generating value of $\mu_\theta^\text{foc}$ is 0.5, but each run will include sampling variability. To get at the heart of how well a method is doing, we calculated the achievement gap residual as the method's estimated achievement gap, $\hat\mu_\theta^\text{foc}$, minus the achievement gap estimated when using only the DIF-free items as anchors. In summary, this outcome measures a method’s ability to disentangle differences in target ability from nuisance ability at the group level.

@@

Assessments are frequently used to make decisions about individual student abilities. We measured a method’s ability to do so by calculating the rank correlation between the vector of estimated abilities and the vector of true target abilities. We used rank correlation as opposed to the more typically used RMSE in agreement with Lord's -@lord1986maximum argument that RMSE can be inflated by poor performance at the extremes of ability for which the test was not designed to measure precisely. *(TODO: this isn't included in this current draft, and I'm not actually sure it's necessary. maybe this idea of rank order being better than RMSE is for another paper)*

>>>>

In our simulation study, it was critical that student ability was drawn in a realistic way similar to @walker2017using. However, we don’t believe that a compensatory model is realistic in describing a math test where some items depend on verbal ability. For example, it’s hard to imagine that a student without the verbal ability to parse a word problem could fully compensate by having a higher math ability. Accordingly, we generated item responses using a simplified version of Sympson’s -@sympson1978model noncompensatory item response model in which
\begin{align}
\text{Pr}(y_{ij} = 1 | \theta_i, \eta_i, a_{2j}) = \sigma(\theta_i) \cdot \sigma(a_{2j}\eta_i)
\end{align}
where, as before, $\theta_i$ is target ability, $\eta_i$ is nuisance ability, and $a_{2j}$ is the item's loading on nuisance ability [@demars2016partially].

@@

The test always has 12 items, but we varied the number of items with DIF from two to six. For items without DIF, $a_{2j} = \infty$ so that the model reduces to $\text{Pr}(y_{ij} = 1 | \theta_i) = \sigma(\theta_i)$. For items with DIF, $a_{2j}$ is calculated based on Ackerman's -@ackerman1994using angle equation as described in @walker2017using:
\begin{align}
\angle_j = \arccos \dfrac{a_{1j}^2}{a_{1j}^2 + a_{2j}^2}.
\end{align}
An item's angle measures the relative loading of the item on the two dimensions. The lower the angle, the more the item loads on target ability as compared to nuisance ability, which corresponds to less DIF. An angle of 45$^\circ$ indicates that the item loads equally on the target ability and nuisance ability. Our simple noncompensatory model has $a_{1j} = 1$ for all items so the angle equation reduces to
\begin{align}
\angle_j = \arccos \dfrac{1}{1 + a_{2j}^2}.
\end{align}
We are interested in specifying the angle of an item, so the relevant equation becomes
\begin{align}
a_{2j} = \sqrt{\dfrac{1 - \cos(\angle_j)^2}{\cos(\angle_j)^2}}.
\end{align}
For DIF items, we set $a_{2j}$ based on angles with equal intervals between 20$^\circ$ and 60$^\circ$. For example, for a test with three DIF items the angles are 20$^\circ$, 40$^\circ$, and 60$^\circ$.

@@

```{r simemmlg, fig.cap = 'For a 12 item test, the relationship between target ability and probability of correct response with nuisance abilities fixed to group means.', out.width="70%", warning = FALSE, message = FALSE}
out$bigsim[[5]]$intuitive_mod[[1]] %>%
    mod_intuitive_to_draws_df() %>%
    draws_df_to_logit_plot() +
    labs(
          x = latex2exp::TeX("$\\tilde{d_j} = \\tilde{b_j}^{ref} - \\tilde{b_j}^{foc}$")
      )
```

@@

<!-- Ben didn't totally foll -->

An effective AI method should lead to a final model that accurately estimates the difference between the reference group's mean target ability and the focal group's mean target ability. We refer to this quantity as the achievement gap. Recall that all models set $\mu_\theta^\text{ref} = 0$, so the achievement gap reduces to $\mu_\theta^\text{foc}$. The data-generating value of $\mu_\theta^\text{foc}$ is 0.5, but each run, of course, includes sampling variability. To get at the heart of how well a method is doing, we calculated the achievement gap residual as the method's estimated achievement gap, $\hat\mu_\theta^\text{foc}$, minus the achievement gap estimated when using only the DIF-free items as anchors. In summary, this outcome measures a method’s ability to disentangle differences in target ability from nuisance ability at the group level.

@@

Assessments are frequently used to make decisions about individual student abilities. We measured a method’s ability to do so by calculating the rank correlation between the vector of estimated abilities and the vector of true target abilities. We used rank correlation as opposed to the more typically used RMSE in agreement with Lord's -@lord1986maximum argument that RMSE can be inflated by poor performance at the extremes of ability for which the test was not designed to measure precisely. *(TODO: this isn't included in this current draft, and I'm not actually sure it's necessary)*
